On the relation of the entanglement spectrum to the bulk polarization

### 2021-04-16 polarization_ES_prb_resubmission/polarization_ES.tex
Add files via upload

One property of the subsystem correlation matrix we will make use of later is that, for sufficiently large systems, its eigenstates are found to be exponentially localized on either virtual edge if $\xi$ is away from $0$ and $1$, or they are found to be bulk modes if the correspondent eigenvalues are exponentially close to $\xi = 0,1$ \cite{Peschel2008}.

@@

In the thermodynamic limit we obtain
\begin{equation}
\lim_{L \rightarrow \infty} \tilde{\mathcal{P}} = \lim_{L \rightarrow \infty} \chi \, {\rm mod } \, 1.
\label{eq:ptilde_eq_chi}
\end{equation}
Note that in order to compute $\chi$ correctly the eigenstates with $\xi \neq 0,1$ need to be localized in order for us to select the correct eigenvalues. It might be --- like in Fig.\ref{huang}(a) --- that the spectrum has degenerate pairs of left and right eigenstates and the numerical diagonalization mixes them. This case is trivial, since we know that there is one left and one right eigenstate, but if the degeneracy is four-fold or higher we must obtain the localized eigenstates. Localizing the bulk eigenstates that are exponentially close to $\xi = 0,1$ is more difficult but not doing so only results in an exponentially small error.

@@

In an attempt to solve these issues we introduce an alternative bulk polarization given by
\begin{equation}\label{eq:tpo}
\tpo = \int_0^{2\pi}\frac{d\Phi}{2\pi} i\brapsio{\Phi}\partial_\Phi \ketpsio{\Phi}.
\end{equation}
where $\ketpsio{{\Phi=0}}$ is the ground state when the (previously periodic) chain is opened between sites $j=L/2$ and $j=L/2+1$ and 
\begin{equation}
\ketpsio{\Phi} =e^{-i \Phi \hat{N}_A} \ketpsio{0}.
\label{eq:gauge}
\end{equation}
When introducing the flux in the system with periodic boundary conditions (See Eq.~\eqref{eq:Ground_state_flux}) there is an ambiguity due to the bulk modes. 
This ambiguity disappears when we introduce the flux as above in the open chain --- we simply include all the bulk modes. 
Note that $\Phi$ is not a magnetic flux but should be treated as an additional parameter, and therefore $\tpo$ is not the polarization of the open chain. $\tpo$ can be obtained in an even simpler way than $\tilde{\mathcal{P}}$ as 
\begin{align}\label{eq:phi_independence}
\tpo =&  \int_0^{2\pi}\frac{d\Phi}{2\pi} \brapsio{0} \hat{N}_A \ketpsio{0} \nonumber \\
=&\sum_{\alpha,j\in A}\brapsio{0} c_{j\alpha}^\dagger c_{j\alpha} \ketpsio{0} \nonumber\\
=&\sum_{\alpha,j\in A} C_{\rm o,j\alpha,j\alpha}\nonumber \\
=& {\rm Tr}[C_{A,\rm o}],
\end{align}
where we used Eq.~\eqref{eq:gauge} and ($C_{A,\rm o}$)$C_{\rm o}$ denotes the (subsystem) correlation matrix of the open chain.

The idea is that in the process of opening the chain, which can be done adiabatically, all eigenvalues that are related to the right virtual edge ($A_R$ in Fig.~\ref{fig:chain}) are pushed to $\xi=0,1$. 
As a result, the sum of all eigenvalues of $C_{A,o}$ (modulo 1) is equal to $\chi$ \eqref{eq:chi_left}. \footnote{ This is strictly speaking only true in the thermodynamic limit, where we avoid the finite size effects due to opening the chain.} Therefore, in the thermodynamic limit, we have
\begin{equation}
\lim_{L \rightarrow \infty }\tilde{\mathcal{P}} = \lim_{L \rightarrow \infty } \tpo \, {\rm mod} \, 1.
\end{equation}

@@

The relation between the zero-modes and discontinuities in $\Tr[C_{A,o}]$ was already observed in the context of Chern insulators \cite{Alexandrinata2011}. The appearance of the branch cut is the result of the degenerate point at $(m=1,\kappa=0)$ having an associated Chern number $C=1$ \cite{Asboth2016}.

@@

We now discuss  the computation of Chern numbers in 2D systems with $\tpo$. Consider the Chern insulator with inversion symmetry given by the Hamiltonian
\begin{align}
H =& \sum_{i\alpha,j\beta}\sum_{k_y} c_{i\alpha}^\dagger(k_y) H_{i\alpha,j\beta}(k_y) c_{j\beta}(k_y), \\ \nonumber
H_{i\alpha,j\beta}(k_y)=& \frac{1}{2}(i\sigma_x-\sigma_z)\delta_{i,j+1}+\frac{1}{2}(-i\sigma_x-\sigma_z)\delta_{i,j-1} \nonumber\\
+&(\sin(k_y)\sigma_y+[2-m-\cos(k_y)]\sigma_z) \delta_{ij}.
\label{eq:chern_model}
\end{align}

@@

\section{Winding number in the SSH chain}
\label{appendix:winding_ssh}

>>>>

One property of the subsystem correlation matrix we will make use of later is that, for sufficiently large systems, its eigenstates are found to be exponentially localized on either virtual edge if $\xi$ is away from $0$ and $1$, or they are found to be bulk modes if the correspondent eigenvalues are exponentially close to $\xi = 0,1$ \cite{Peschel2008}. The latter can be understood from the fact that the eigenstates of $C$ can be expressed as Wannier states. The Wannier states for the occupied ($\xi=1$) or empty ($\xi=0$) bands that are contained within region $A$ will also be eigenstates of $C_A$, with only a small deviation due to their exponential tails. However, in practice, all these eigenstates mix at $\xi=0,1$, forming the two bulk bands.

@@

In the thermodynamic limit we obtain
\begin{equation}
\lim_{L \rightarrow \infty} \tilde{\mathcal{P}} = \lim_{L \rightarrow \infty} \chi \, {\rm mod } \, 1.
\label{eq:ptilde_eq_chi}
\end{equation}
Note that in order to compute $\chi$ correctly the eigenstates with $\xi \neq 0,1$ need to be localized in order for us to select the correct eigenvalues. It might be --- like in Fig.\ref{huang}(a) --- that the spectrum has degenerate pairs of left and right eigenstates and the numerical diagonalization mixes them. This case is trivial, since we know that there is one left and one right eigenstate, but if the degeneracy is four-fold or higher we must obtain the localized eigenstates. One way of counting how many left eigenstates are in a degenerate subspace is discussed in Appendix \ref{app:loc}. Note that this is not necessary for the bulk eigenstates, exponentially close to $\chi=0,1$, as the error we make is also exponentially small once we take the result modulo 1.

@@

In an attempt to solve these issues we introduce an alternative bulk polarization given by
\begin{equation}\label{eq:tpo}
\tpo = \int_0^{2\pi}\frac{d\Phi}{2\pi} i\brapsio{\Phi}\partial_\Phi \ketpsio{\Phi}.
\end{equation}
where $\ketpsio{{\Phi=0}}$ is the ground state when the (previously periodic) chain is opened between sites $j=L/2$ and $j=L/2+1$ and 
\begin{equation}
\ketpsio{\Phi} =e^{-i \Phi \hat{N}_A} \ketpsio{0},
\label{eq:gauge}
\end{equation}
where $\hat{N}_A$ is the number operator for particles in region $A$.

When introducing the flux in the system with periodic boundary conditions (See Eq.~\eqref{eq:Ground_state_flux}) there is an ambiguity due to the bulk modes. 
This ambiguity disappears when we introduce the flux as above in the open chain --- we simply include all the bulk modes. 
$\tpo$ can be obtained in an even simpler way than $\tilde{\mathcal{P}}$ as 
\begin{align}\label{eq:phi_independence}
\tpo =&  \int_0^{2\pi}\frac{d\Phi}{2\pi} \brapsio{0} \hat{N}_A \ketpsio{0} \nonumber \\
=&\sum_{\alpha,j\in A}\brapsio{0} c_{j\alpha}^\dagger c_{j\alpha} \ketpsio{0} \nonumber\\
=&\sum_{\alpha,j\in A} C_{\rm o,j\alpha,j\alpha}\nonumber \\
=& {\rm Tr}[C_{A,\rm o}],
\end{align}
where we used Eq.~\eqref{eq:gauge} and ($C_{A,\rm o}$)$C_{\rm o}$ denotes the (subsystem) correlation matrix of the open chain.

Note that the result is very easy to interpret as well. Changes in the polarization $\tilde{\mathcal{P}}$ on the periodic chain give the current across the seam. In the open chain, this is simply given by $J = \partial_t N_A(t)$. Therefore the equivalent polarization in the open chain is simply the number of particles in region $A$, $N_A$. It is, however, useful to define it in terms of the flux, as we are interested in studying the polarization as a geometric quantity. This description is more subtle when states cross the chemical potential, where $N_A$ will have a discontinuity without an associated current. Discontinuities in the trace of $C_A$ for open chains was already observed in the context of Chern insulators \cite{Alexandrinata2011}, where it was related to the appearance of gapless modes. Below we will employ similar arguments to study the topology of 1D chains.

The key idea to understand the connection of the EOS for open and periodic chains is that in the process of opening the chain, which can be done adiabatically, all eigenvalues that are related to the right virtual edge ($A_R$ in Fig.~\ref{fig:chain}) are pushed to $\xi=0,1$. 
As a result, the sum of all eigenvalues of $C_{A,o}$ (modulo 1) is equal to $\chi$ \eqref{eq:chi_left}. \footnote{ This is strictly speaking only true in the thermodynamic limit, where we avoid the finite size effects due to opening the chain.} Therefore, in the thermodynamic limit, we have
\begin{equation}
\lim_{L \rightarrow \infty }\tilde{\mathcal{P}} = \lim_{L \rightarrow \infty } \tpo \, {\rm mod} \, 1.
\end{equation}

@@

The appearance of the branch cut is the result of the degenerate point at $(m=1,\kappa=0)$ having an associated Chern number $C=1$ \cite{Asboth2016}.

@@

We now discuss the computation of Chern numbers in 2D systems with $\tpo$ by using the same method we described for 1D systems. A similar (related?) discussion can be found in ref.\cite{Alexandrinata2011}.
 Consider the Chern insulator with inversion symmetry given by the Hamiltonian
\begin{align}
H =& \sum_{i\alpha,j\beta}\sum_{k_y} c_{i\alpha}^\dagger(k_y) H_{i\alpha,j\beta}(k_y) c_{j\beta}(k_y), \\ \nonumber
H_{i\alpha,j\beta}(k_y)=& \frac{1}{2}(i\sigma_x-\sigma_z)\delta_{i,j+1}+\frac{1}{2}(-i\sigma_x-\sigma_z)\delta_{i,j-1} \nonumber\\
+&(\sin(k_y)\sigma_y+[2-m-\cos(k_y)]\sigma_z) \delta_{ij}.
\label{eq:chern_model}
\end{align}

@@

\section{Degenerate EOS}\label{app:loc}

In order apply Eq.\eqref{eq:chi_left} one needs to identify to which virtual cut each eigenstate belongs to. In the case where the EOS is degenerate, eigenstates from the two different virtual cuts might mix, making this identification not straightforward . One solution to this problem is as follows. Consider the subspace spanned by the degenerate eigenstates $\ket{\psi_\mu}$. These are a mixture of states $\ket{\tilde{\psi}_\nu}$ that are localized either on the left or on the right half of region $A$. Consider the projector onto this subspace,
\begin{align}
P_{deg} =& \sum_\mu \ket{\psi_\mu}\bra{\psi_{\mu}} \nonumber \\
=&  \sum_{\nu \in L} \ket{\tilde{\psi}_\mu}\bra{\tilde{\psi}_{\mu}}+ \sum_{\nu \in R} \ket{\tilde{\psi}_\mu}\bra{\tilde{\psi}_{\mu}} 
\end{align}
We want to count the number of these localized eigenstates in region $L$. This can be done simply by taking the trace over region $L$, such that
\begin{align}
\Tr_L [P_{deg} ]=&  \Tr_L[ \sum_{\nu \in L} \ket{\tilde{\psi}_\mu}\bra{\tilde{\psi}_{\mu}}]+ \Tr_L[ \sum_{\nu \in R} \ket{\tilde{\psi}_\mu}\bra{\tilde{\psi}_{\mu}} ] \\
=&  \sum_{\nu \in L} \bra{\tilde{\psi}_{\mu}} \ket{\tilde{\psi}_\mu} .
\end{align}
\section{Winding number in the SSH chain}
\label{appendix:winding_ssh}
